Static Analysis by Path Focusing

### 2011-06-09 report/report.tex
report update

\tikzstyle{state}=[circle,fill=black!25,minimum size=13pt,inner sep=0pt]
\tikzstyle{transition}=[rectangle,semithick,draw=black!75,
  			  minimum size=4mm]
\tikzstyle{transition2}=[transition,rectangle,thick,dashed,
  			  minimum size=4mm]
\tikzstyle{PRstate}=[circle,double,draw,fill=blue!15,minimum size=13pt,inner sep=0pt]
\tikzstyle{polyhedra}=[blue!25,opacity=0.5,pattern=north west lines,pattern
color=blue]
\tikzstyle{line}=[black,thick]

@@

\title{Static Analysis by Path Focusing}
\author{Julien Henry}
\date{2011}
\shorttitle{Static Analysis by Path Focusing}
\institute{Grenoble-INP}
\abstract{
	Program verification is aims at statically discovering properties on programs,
	such as the values that can take the different variables during execution.
	Abstract Interpretation is a technique that allows to compute an
	approximation of the set of these values, since it is impossible to compute
	the real set in general.
	This report takes place in the many attempts to improve the precision of
	static Analysis by Abstract Interpretation. It proposes a technique that
	takes benefit of SMT-solving to obtain more precise results at reasonable
	cost.
}
\resume{  
	La vérification de programme consiste à découvrir statiquement des
	propriétés sur des programmes, comme l'ensemble des valeurs que peuvent
	prendre les variables durant l'exécution. L'Interprétation Abstraite est une
	technique permettant de calculer une approximation de cet ensemble, le
	véritable ensemble étant impossible à calculer en général.
	Ce report s'inscrit dans la lignée des travaux visant à améliorer la
	précision de l'analyse par interprétation abstraite, et propose une
	technique tirant parti du SMT-solving pour obtenir de meilleurs résultats
	à un coût raisonnable.
}
\keywords{Static Program Analysis, Verification, Abstract Interpretation,
Invariant Generation, Path Focusing, SMT-Solving}
\motscles{Analyse Statique de Programmes, Vérification, Interprétation
Abstraite, Génération d'invariants, Découverte de Chemins, SMT-Solving}
\tutor{David Monniaux - Matthieu Moy}
\notes{some notes}
\webaddress{Julien.Henry@imag.fr}
\reporttype{M2R internship's report}

@@

\begin{algorithm}
\caption{SelfLoop}
\label{loopiteralgo}
\begin{algorithmic}[1] 
\Procedure{SelfLoop}{$\tau_e^\#$,$X_{p_i}$}
	\State $Y \gets \tau_e^\#(X_{p_i})$
	\State $X' \gets X_{p_i} \sqcup Y$
	\State $Y \gets \tau_e^\#(X')$
	\State \textbf{return} $Y$
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

	We apply the \emph{Path Focusing} technique to our running example from
	Figures \ref{runningexample}, \ref{multigraphtransformGopan} and
	\ref{multigraphGopan}.
	The control points in $P_R$ are $p_0$, $p_1$ and $p_7$. 
	The $\rho$ formula associated to this program is depicted in
	Figure \ref{rhoformula}.
	One computes
	iteratively the associated abstract values $X_0$, $X_1$ and $X_7$.

@@

\begin{table}[!h]
\centering
\begin{tabular}{|c||c|c|c|c|} \hline 
Benchmark & Path Focusing & Lookahead Widening & Same invariant & Total \\
\hline \hline
automotive\_bitcount	& 1 & 0 & 39 & 40 \\ \hline
automotive\_qsort1		& 7 & 0 & 9 & 16 \\ \hline
automotive\_susan\_c	& 0 & 0 & 3 & 3 \\ \hline
automotive\_susan\_e	& 0 & 0 & 3 & 3 \\ \hline
automotive\_susan\_s	& 0 & 0 & 3 & 3 \\ \hline
bzip2d					& 8 & 0 & 22 & 30 \\ \hline
bzip2e					& 8 & 0 & 22 & 30 \\ \hline
consumer\_jpeg\_c		& 35 & 0 & 693 & 728 \\ \hline
consumer\_jpeg\_d		& 30 & 0 & 662 & 692 \\ \hline
consumer\_lame			& 31 & 12 & 372  & 415 \\ \hline
consumer\_mad			& 9 & 8 & 579 & 596 \\ \hline
consumer\_tiff2bw		& 18 & 1 & 576 & 595 \\ \hline
consumer\_tiff2rgba		& 19 & 1 & 574 & 594\\ \hline
consumer\_tiffdither	& 17 & 1 & 574 & 592 \\ \hline
consumer\_tiffmedian	& 21 & 2 & 620 & 643 \\ \hline
network\_patricia		& 2 & 0 & 21 & 23 \\ \hline
office\_rsynth			& 8 & 0 & 158 & 166\\ \hline
office\_stringsearch1	& 5 & 1 & 44 & 50\\ \hline
security\_blowfish\_d	& 4 & 0 & 23 & 27\\ \hline
security\_blowfish\_e	& 4 & 0 & 23 & 27\\ \hline
security\_rijndael\_d	& 2 & 0 & 30 & 32\\ \hline
security\_rijndael\_e	& 2 & 0 & 30 & 32\\ \hline
security\_sha			& 1 & 0 & 26 & 27\\ \hline
telecom\_adpcm\_c		& 0 & 0 & 10 & 10\\ \hline
telecom\_adpcm\_d		& 0 & 0 & 10 & 10\\ \hline
telecom\_CRC32			& 0 & 0 & 13 & 13\\ \hline
telecom\_gsm			& 4 & 0 & 89 & 93\\ \hline
\end{tabular}
\caption{Number of basicblocks where the computed invariant is more precise by
\emph{Path Focusing}, \emph{Lookahead Widening}, or the same with both
techniques.}
\label{cbench}
\end{table}

@@

\begin{table}
\centering
\begin{tabular}{|c||c|c|c|c|} \hline 
Benchmark & Path Focusing & Lookahead Widening & Same invariant & Total \\
\hline \hline
Bc			& 0 & 0 & 177 & 177 \\ \hline
Gawk		& 4 & 0 & 287 & 291 \\ \hline
Gnuchess	& 16 & 16 & 1447 & 1479 \\ \hline
Gnugo		& 120 & 10 & 1286 & 1416 \\ \hline
Grep		& 4 & 1 & 326 & 331 \\ \hline
Gzip		& 16 & 8 & 207 & 231 \\ \hline
Make		&  5 & 5 & 304 & 314 \\ \hline
Tar			& 13 & 2 & 437 & 452 \\ \hline
Wget		& 20 & 3 & 751 & 774 \\ \hline
\end{tabular}
\caption{Result of the analysis of various open-source projects.}
\label{opensource}
\end{table}

@@

\addcontentsline{toc}{chapter}{Bibliography}
  \appendix
\begin{figure}
\begin{C}
(and (= bd_0 false)
        (= t_0_1 bs_0)
        (or (not bd_0) true)
        (= b_5 t_1_5)
        (= x_%6_ (+ x_y.0_ -1))
        (= t_5_7 b_5)
        (or (not b_5) true)
        (= b_3 t_1_3)
        (= x_%4_ (+ x_y.0_ 1))
        (= t_3_7 b_3)
        (or (not b_3) true)
        (= b_7
           (or t_5_7 t_3_7))
        (= t_7_9 (and b_7 (< x_y.1_ 0)))
        (= t_7_10
           (and b_7 (not (< x_y.1_ 0))))
        (or (not b_7)
            (= x_y.1_ (ite t_3_7 x_%4_ x_%6_)))
        (= b_10 t_7_10)
        (= x_%11_ (+ x_x.0_ 1))
        (= t_10_1 b_10)
        (or (not b_10) true)
        (= bd_1
           (or t_10_1 t_0_1))
        (= t_1_3 (and bs_1 (<= x_x.0_ 50)))
        (= t_1_5
           (and bs_1 (not (<= x_x.0_ 50))))
        (or (not bd_1)
            (and (= x'_x.0_ (ite t_0_1 0 x_%11_))
                 (= x'_y.0_ (ite t_0_1 0 x_y.1_))))
        (= b_9 t_7_9)
        (= t_9_12 b_9)
        (or (not b_9) true)
        (= bd_12 t_9_12)
        (or (not bd_12) true))
\end{C}
\caption{Expression of $\rho$ associated to the program in Figures
\ref{runningexample}, \ref{multigraphtransformGopan} and
\ref{multigraphGopan}, in the SMT-lib format.}
\label{rhoformula}
\end{figure}

>>>>

\tikzstyle{state}=[circle,fill=black!25,minimum size=13pt,inner sep=0pt]
\tikzstyle{rstate}=[rectangle,fill=black!25,minimum size=13pt,inner sep=0pt]
\tikzstyle{transition}=[rectangle,semithick,draw=black!75,
  			  minimum size=4mm]
\tikzstyle{transition2}=[transition,rectangle,thick,dashed,
  			  minimum size=4mm]
\tikzstyle{PRstate}=[circle,double,draw,fill=blue!15,minimum size=13pt,inner sep=0pt]
\tikzstyle{polyhedra}=[blue!25,opacity=0.5,pattern=north west lines,pattern
color=blue]
\tikzstyle{line}=[black,thick]

@@

\lstnewenvironment{Csmall}
{\lstset{language=C,
		basicstyle=\ttfamily\small,
		commentstyle=\color{moka}\textit,
		keywordstyle=\color{minuit},
		identifierstyle=\color{trefle},
		showstringspaces=false}}
{}
\title{Static Analysis by Path Focusing}
\author{Julien Henry}
\date{2011}
\shorttitle{Static Analysis by Path Focusing}
\institute{Grenoble-INP}
\abstract{
	Program verification is aims at statically discovering properties on programs,
	such as the values that can take the different variables during execution.
	Abstract Interpretation is a technique that allows to compute an
	approximation of the set of these values, since it is impossible to compute
	the real set in general.
	This report takes place in the many attempts to improve the precision of
	static Analysis by Abstract Interpretation. It proposes a technique that
	takes benefit of SMT-solving to obtain more precise results at reasonable
	cost.
}
\resume{  
	La vérification de programme consiste à découvrir statiquement des
	propriétés sur des programmes, comme l'ensemble des valeurs que peuvent
	prendre les variables durant l'exécution. L'Interprétation Abstraite est une
	technique permettant de calculer une approximation de cet ensemble, le
	véritable ensemble étant impossible à calculer en général.
	Ce report s'inscrit dans la lignée des travaux visant à améliorer la
	précision de l'analyse par interprétation abstraite, et propose une
	technique tirant parti du SMT-solving pour obtenir de meilleurs résultats
	à un coût raisonnable.
}
\keywords{Static Program Analysis, Verification, Abstract Interpretation,
Invariant Generation, Path Focusing, SMT-Solving}
\motscles{Analyse Statique de Programmes, Vérification, Interprétation
Abstraite, Génération d'invariants, Découverte de Chemins, SMT-Solving}
\tutor{David Monniaux - Matthieu Moy}
%\notes{some notes}
\webaddress{Julien.Henry@imag.fr}
\reporttype{M2R internship's report}

\acknowledgement{
	I would like to thank my two tutors, David Monniaux and Matthieu Moy, for
	their direction of my master thesis, and for their help during all my
	internship. I thank them for all their explanations that clearified lots of
	my questions, and for their advices for the redaction of this report.
}

@@

\begin{algorithm}
\caption{SelfLoop}
\label{loopiteralgo}
\begin{algorithmic}[1] 
\Procedure{SelfLoop}{$\tau_e^\#$,$X_{p_i}$}
	\State $Y \gets \tau_e^\#(X_{p_i})$
	\State $X' \gets X_{p_i}\widening (X_{p_i} \sqcup Y)$
	\State $Y \gets \tau_e^\#(X')$
	\State \textbf{return} $Y$
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

	We apply the \emph{Path Focusing} technique to our running example from
	Figures \ref{runningexample}, \ref{multigraphtransformGopan} and
	\ref{multigraphGopan}.
	The control points in $P_R$ are $p_0$, $p_1$ and $p_7$. 
	The $\rho$ formula associated to this program is depicted in
	Figure \ref{rhoformula} (see appendices).
	One computes
	iteratively the associated abstract values $X_0$, $X_1$ and $X_7$.

@@

\begin{table}[!h]
\centering
\begin{tabular}{|c||c|c|c|c|} \hline 
Benchmark & Path Focusing & Lookahead Widening & Same invariant & Total \\
\hline \hline
%automotive\_bitcount	& 1 & 0 & 39 & 40 \\ \hline
%automotive\_qsort1		& 7 & 0 & 9 & 16 \\ \hline
%automotive\_susan\_c	& 0 & 0 & 3 & 3 \\ \hline
%automotive\_susan\_e	& 0 & 0 & 3 & 3 \\ \hline
%automotive\_susan\_s	& 0 & 0 & 3 & 3 \\ \hline
%bzip2d					& 8 & 0 & 22 & 30 \\ \hline
%bzip2e					& 8 & 0 & 22 & 30 \\ \hline
%consumer\_jpeg\_c		& 35 & 0 & 693 & 728 \\ \hline
%consumer\_jpeg\_d		& 30 & 0 & 662 & 692 \\ \hline
%consumer\_lame			& 31 & 12 & 372  & 415 \\ \hline
%consumer\_mad			& 9 & 8 & 579 & 596 \\ \hline
%consumer\_tiff2bw		& 18 & 1 & 576 & 595 \\ \hline
%consumer\_tiff2rgba		& 19 & 1 & 574 & 594\\ \hline
%consumer\_tiffdither	& 17 & 1 & 574 & 592 \\ \hline
%consumer\_tiffmedian	& 21 & 2 & 620 & 643 \\ \hline
%network\_patricia		& 2 & 0 & 21 & 23 \\ \hline
%office\_rsynth			& 8 & 0 & 158 & 166\\ \hline
%office\_stringsearch1	& 5 & 1 & 44 & 50\\ \hline
%security\_blowfish\_d	& 4 & 0 & 23 & 27\\ \hline
%security\_blowfish\_e	& 4 & 0 & 23 & 27\\ \hline
%security\_rijndael\_d	& 2 & 0 & 30 & 32\\ \hline
%security\_rijndael\_e	& 2 & 0 & 30 & 32\\ \hline
%security\_sha			& 1 & 0 & 26 & 27\\ \hline
%telecom\_adpcm\_c		& 0 & 0 & 10 & 10\\ \hline
%telecom\_adpcm\_d		& 0 & 0 & 10 & 10\\ \hline
%telecom\_CRC32			& 0 & 0 & 13 & 13\\ \hline
%telecom\_gsm			& 4 & 0 & 89 & 93\\ \hline
%%%%%%%%%%%%
automotive\_bitcount & 1 & 0 & 39 & 40 \\ \hline
automotive\_qsort1 & 9 & 0 & 13 & 22 \\ \hline
automotive\_susan\_c & 0 & 2 & 59 & 61 \\ \hline
automotive\_susan\_e & 0 & 2 & 59 & 61 \\ \hline
automotive\_susan\_s & 0 & 2 & 59 & 61 \\ \hline
bzip2d & 70 & 3 & 81 & 154 \\ \hline
bzip2e & 70 & 3 & 81 & 154 \\ \hline
consumer\_jpeg\_c & 22 & 8 & 711 & 741 \\ \hline
consumer\_jpeg\_d & 13 & 8 & 667 & 688 \\ \hline
consumer\_lame & 13 & 15 & 366 & 394 \\ \hline
consumer\_mad & 12 & 11 & 609 & 632 \\ \hline
consumer\_tiff2bw & 19 & 3 & 593 & 615 \\ \hline
consumer\_tiff2rgba & 21 & 6 & 661 & 688 \\ \hline
consumer\_tiffdither & 21 & 3 & 662 & 686 \\ \hline
consumer\_tiffmedian & 25 & 5 & 707 & 737 \\ \hline
network\_dijkstra & 0 & 0 & 19 & 19 \\ \hline
network\_patricia & 2 & 0 & 21 & 23 \\ \hline
office\_rsynth & 6 & 0 & 158 & 164 \\ \hline
office\_stringsearch1 & 5 & 1 & 44 & 50 \\ \hline
security\_blowfish\_d & 4 & 0 & 23 & 27 \\ \hline
security\_blowfish\_e & 4 & 0 & 23 & 27 \\ \hline
security\_rijndael\_d & 1 & 0 & 31 & 32 \\ \hline
security\_rijndael\_e & 1 & 0 & 31 & 32 \\ \hline
security\_sha & 1 & 0 & 26 & 27 \\ \hline
telecom\_CRC32 & 0 & 0 & 13 & 13 \\ \hline
telecom\_adpcm\_c & 0 & 0 & 10 & 10 \\ \hline
telecom\_adpcm\_d & 0 & 0 & 10 & 10 \\ \hline
telecom\_gsm & 5 & 0 & 153 & 158 \\ \hline
\end{tabular}
\caption{Number of basicblocks where the computed invariant is more precise by
\emph{Path Focusing}, \emph{Lookahead Widening}, or the same with both
techniques.}
\label{cbench}
\end{table}

@@

The results we obtain are very promising. Most of the cases, our technique
performs as well as \emph{Lookahead Widening}, and discovers regularly more
precise invariants.

However, \emph{Lookahead Widening} seems to have better results in some cases.
We already have some ideas for explaining this lack of precision in our
technique:

\begin{itemize}
\item We use the domain of convex polyhedra as the abstract domain. Since this
domain requires a widening operator, a more subtle analysis technique may
sometimes discover invariants that are less precise than a more simplified
technique.
\item Our technique performs well in the case of self loops in the multigraph,
thanks to the widening/narrowing phases (or acceleration). Yet, in the case of
paths that do not constitute a self loop in the multigraph, one perform a
widening operation, but no narrowing iteration. Apparently, lots of the cases
where \emph{Lookahead Widening} performs better than \emph{Path Focusing} are
due to this drawback. An example of such loss of precision is depicted in
appendix \ref{lostprecision}.
An good idea that could solve this problem is to try to combine
the two methods.
\end{itemize}

Although these project are not safety-critical and do not necessarily need
static analysis, they give a good idea of the quality of our technique and the
scalability of our implementation.

\begin{table}
\centering
\begin{tabular}{|c||c|c|c|c|} \hline 
Benchmark & Path Focusing & Lookahead Widening & Same invariant & Total \\
\hline \hline
Bc			& 0 & 0 & 177 & 177 \\ \hline
Gawk		& 4 & 3 & 284 & 291 \\ \hline
Gnuchess	& 22 & 33 & 1506 & 1561 \\ \hline
Gnugo		& 105 & 35 & 1303 & 1443 \\ \hline
Grep		& 4 & 3 & 323 & 330 \\ \hline
Gzip		& 9 & 3 & 189 & 201 \\ \hline
Make		&  7 & 11 & 457 & 475 \\ \hline
Tar			& 16 & 5 & 555 & 576 \\ \hline
Wget		& 8 & 12 & 715 & 735 \\ \hline
\end{tabular}
\caption{Result of the analysis of various open-source projects.}
\label{opensource}
\end{table}

@@

\addcontentsline{toc}{chapter}{Bibliography}
  \appendix
\appendixpage
\renewcommand{\thesection}{\Alph{section}}
\section{Generated $\rho$ formula}
\label{rhoformula}
\begin{figure}[!h]
\begin{C}
(and (= bd_0 false)
        (= t_0_1 bs_0)
        (or (not bd_0) true)
        (= b_5 t_1_5)
        (= x_%6_ (+ x_y.0_ -1))
        (= t_5_7 b_5)
        (or (not b_5) true)
        (= b_3 t_1_3)
        (= x_%4_ (+ x_y.0_ 1))
        (= t_3_7 b_3)
        (or (not b_3) true)
        (= b_7
           (or t_5_7 t_3_7))
        (= t_7_9 (and b_7 (< x_y.1_ 0)))
        (= t_7_10
           (and b_7 (not (< x_y.1_ 0))))
        (or (not b_7)
            (= x_y.1_ (ite t_3_7 x_%4_ x_%6_)))
        (= b_10 t_7_10)
        (= x_%11_ (+ x_x.0_ 1))
        (= t_10_1 b_10)
        (or (not b_10) true)
        (= bd_1
           (or t_10_1 t_0_1))
        (= t_1_3 (and bs_1 (<= x_x.0_ 50)))
        (= t_1_5
           (and bs_1 (not (<= x_x.0_ 50))))
        (or (not bd_1)
            (and (= x'_x.0_ (ite t_0_1 0 x_%11_))
                 (= x'_y.0_ (ite t_0_1 0 x_y.1_))))
        (= b_9 t_7_9)
        (= t_9_12 b_9)
        (or (not b_9) true)
        (= bd_12 t_9_12)
        (or (not bd_12) true))
\end{C}
\caption{Expression of $\rho$ associated to the program in Figures
\ref{runningexample}, \ref{multigraphtransformGopan} and
\ref{multigraphGopan}, in the SMT-lib format.}
\end{figure}
\FloatBarrier

\newpage
\section{Example of loss of precision in Path Focusing}
\label{lostprecision}
The following graph is a simplified example taken from one of our benchmarks,
where \emph{Path Focusing} technique obtains bad results.
\begin{figure}[!h]
\centering
\begin{tikzpicture}[->,>=stealth',auto,node distance=2.5cm,
                    semithick,font=\footnotesize]

	\node[rstate] (n0) {$p_0$};
	\node[rstate] (n1) [below of=n0] {\begin{tabular}{l}
	$p_1$: \\
	$i \gets \Phi(0,i')$
	\end{tabular}};
	\node[rstate] (n2) [below left of=n1] {$p_2$};
	\node[rstate] (n3) [below right of=n2] {$p_3$};

  \path [transition] 
		(n0) edge              node {} (n1);
  \path [transition] 
        (n1) edge			   node [left] {$i \leq 50$} (n2);
  \path [transition] 
        (n2) edge	[loop left, distance=2cm]	   node  {} (n2);
  \path [transition] 
        (n2) edge			   node  {} (n3);
  \path [transition] 
        (n3) edge			   node [right] {$i' \gets i+1$} (n1);
\end{tikzpicture}
\end{figure}

$p_1$ and $p_2$ are both loop headers. We have $P_R = \{p_1, p_2\}$.

\begin{itemize}
\item We start with $p_1$ : $i=0$. The SMT-solver finds the path $p_1 \rightarrow p_2$.
\item At point $p_2$ : $i=0$. Then, we focus on the path $p_2 \rightarrow p_3 \rightarrow p_1$.
The obtained invariant in $p_1$ becomes $i \in [0,1]$. 
Since this is a loop header, widening is applied: $i \geq 0$.
\item The new focus path is $p_1 \rightarrow p_2$. The image of the polyhedron $i
\geq 0$ by this transition is $0 \leq i \leq 50$. Again, $p_2$  is a loop
header, and widening gives $i \geq 0$.
\end{itemize}
In this example, we do not have narrowing iterations since we never focus on self
loops.
